---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: Referencias.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(deltapif)
```

## Introduction

The `deltapif` package calculates **Potential Impact Fractions (PIF) and Population Attributable Fractions (PAF)** for summary data, along with their confidence intervals, using the delta method. This vignette provides a practical introduction to the package's core functionality.

### Core Concepts: PAF and PIF

The **Population Attributable Fraction (PAF)** answers the question: _"What fraction of disease cases in a population would be prevented if we completely eliminated a risk factor?"_ It represents the maximum possible reduction achievable and is often interpreted as the burden of disease attributed to the exposure.

The **Potential Impact Fraction (PIF)** is a more general measure. It answers: _"What fraction of disease cases would be prevented if we changed exposure from its current distribution to a specific counterfactual scenario?"_ 

PAF is a specific type of PIF where the counterfactual scenario is the **theoretical minimum risk exposure level (TMREL)**. We remark that the TMREL is not always zero. For example, the TMREL for sodium intake is a specific healthy range (e.g., ~1.6g/day), as both too much and too little sodium are harmful. For other exposures, such as smoking, the TMREL can indeed be zero exposure.

Example:

+ @kathy estimates the reduction in cardiovascular disease cases if the sodium content in packaged foods was reduced to meet Australian's government targets. This is a PIF.

+ Estimating the reduction if sodium consumption was changed to the ideal TMREL of 1.6g/day would be a PAF, representing the maximum achievable benefit.

### Why the Delta Method?

This package uses the delta method to calculate confidence intervals for PAF and PIF. This approach offers key advantages:

+ **Distribution-Free:** It doesn't require assumptions about the underlying distribution of the exposure data in the population, which can be difficult to verify and lead to bias if misspecified.

+ **Computational Efficiency:**It provides a direct algebraic formula for the variance, making it faster and more stable than simulation-based methods like Monte Carlo.

+ **Transparency:** The formula can be derived and understood, offering clarity over "black box" methods.

### Key assumption: Independent (summary) data sources. 

The `deltapif` package is designed for a specific, common scenario in public health:

+ The estimate of the relative risk (`beta`) comes from one source (e.g., a published meta-analysis).

+ The estimate of the exposure prevalence (`p`) comes from a separate, independent source (e.g., a national survey).

The delta method implementation here relies on this independence. If you have individual-level data for exposure the [`pifpaf` package](https://github.com/INSP-RH/pifpaf) is more appropriate as it leverages the individual-level variability. If individual-level exposure and outcome data is available from the same source the [`graphPAF` package](https://cran.r-project.org/web/packages/graphPAF/index.html) is ideal. 

## Examples

### Example 1: Estimating the PAF for the proportion of dementia attributable to smoking

The article by [@lee](https://doi.org/10.1001/jamanetworkopen.2022.19672) estimates the PAF of dementia associated with 12 risk factors (`logrr`) in US adults. The package includes this data as well as the proportion of exposed individuals in the respective race column (Hispanic, Asian, Black and White):

```{r}
data(dementiarisk)
```


```{r, echo = FALSE}
dementiarisk[,c("risk_factor","logrr","sdlog", "total", "hispanic", "asian", "black", "white")]
```

In this example, we show how to calculate the population attributable fraction for **Smoking** among the 4 race groups and then how to calculate the **Total Population Attributable Fraction** of the overall population. 


Notice that **Smoking** has a log relative risk of:

+ `beta = 0.4637340` 

with variance of	

+ `var_beta = 0.0273858` (`= 0.165486566^2`). 

Among hispanic individuals 6.9% (`p = 0.069`) smoke. Hence their PAF is:

```{r}
paf_hispanic <- paf(p = 0.069, beta = 0.4637340, var_beta = 0.0273858, var_p = 0,
                    rr_link = exp, label = "Hispanic")
paf_hispanic
```

The other fractions can be computed in the same way noting that 4.9%	of non-hispanic asians, 11.7% of non-hispanic blacks and	8.4% of non-hispanic whites smoke:

```{r}
paf_asian <- paf(p = 0.049, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "Asian")
paf_black <- paf(p = 0.117, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "Black")
paf_white <- paf(p = 0.084, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "White")
```


#### Combining subpopulations into a total

We can calculate the **total population attributable fraction** by combining the fractions of the subpopulations weighted by the proportion of the population. According to Wikipedia the distribution in 2020 of the US population by race was as follows:

```{r}
weights <- c("white" = 0.5784, "hispanic" = 0.1873, 
             "black" = 0.1205, "asian" = 0.0592)

#Normalized weights to sum to 1
weights <- weights / sum(weights)
```

The `paf_total` function computes the aggregated `paf` of the whole population:

```{r}
paf_population <- paf_total(paf_white, paf_hispanic, paf_black, 
                            paf_asian, weights = weights, var_weights = 0,
                            label = "All")
paf_population
```

where we set `var_weights = 0` as no covariance matrix was known for the race distribution data (it does exist its just not on Wikipedia). 

The `paf` total implements the following formula:

$$
\textrm{PAF}_{\text{Total}} = w_1 \cdot \text{PAF}_{\text{White}} + 
  w_2 \cdot \text{PAF}_{\text{Hispanic}} +
  w_3 \cdot \text{PAF}_{\text{Black}} +
  w_4 \cdot \text{PAF}_{\text{Asian}}
$$

where the weights stand for the proportion of the population in each subgroup. 

The `coef`, `confint`, `summary`, and `as.data.frame` functions have been extended to 
allow the user to extract information from a `paf` (or `pif`):

```{r}
as.data.frame(paf_population)
```

Attributable cases can be calculated for example using the estimates from 
[@dhana2023prevalence](https://pmc.ncbi.nlm.nih.gov/articles/PMC10593099/#SD2). They estimate the number of people with Alzheimer's Disease in New York, USA 426.5 (400.2, 452.7) thousand. This implies a variance of `((452.7 - 400.2) / 2*qnorm(0.975))^2 = 2647.005`. 

```{r}
attributable_cases(426.5, paf_population, variance = 2647)
```

#### Why are we using the log relative risk?

A common question is why in this case we are using the log relative risk (`logrr`) with `exp` link and not the relative risks directly. The reason depends on what standard deviation can be properly
recovered. Recall that the classic (Wald-type) confidence intervals should have the same
distance from the point estimate (in this case the `RR`) to the bounds. We can see that
this does not happen here as for example in the case of **smoking** the RR is
1.59 [1.15-2.20] and:

```{r}
2.20 - 1.59
```

while 

```{r}
1.59 - 1.15
```

The distances are not the same!

However in the log-scale they are much more similar:

```{r}
log(2.20) - log(1.59)
```

```{r}
log(1.59) - log(1.15)
```


which suggests that the uncertainty was estimated in the log-scale and hence we
need to use the log relative risk as beta and exponentiate it (`link = exp`). 
Furthermore we can go to the original paper, check the methodology and conclude
that the confidence interval (and hence the variance!) was estimated using the log-relative risk 
and then exponentiated. Hence  the uncertainty should be propagated for the 
log risk and then exponentiated. 

Mathematically the explanation of why this happens is because usually relative risks,
odds ratios and hazard ratios are transformed to the log-scale to estimate
the confidence intervals. You can see in Wikipedia, for example, 
the classic formulas for the [relative risks CIs](https://en.wikipedia.org/wiki/Relative_risk)
and for the [odds ratios](https://en.wikipedia.org/wiki/Odds_ratio#Statistical_inference) all involve a log-transform. 

### Example 2: Estimating the PIF for the proportion of dementia reduced by a 15% reduction in smoking. 

[@lee](https://doi.org/10.1001/jamanetworkopen.2022.19672) also estimates the proportion of dementia cases that would be reduced by a 15% decrease in smoking prevalence. This 15% reduction corresponds to multiplying the current smoking prevalence by `1 - 0.15 = 0.85`  specified with the prevalence under the counterfactual (`p_cft`) variable: 

```{r}
pif_hispanic <- pif(p = 0.069, p_cft = 0.069*0.85, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "Hispanic")
pif_asian    <- pif(p = 0.049, p_cft = 0.049*0.85, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "Asian")
pif_black    <- pif(p = 0.117, p_cft = 0.117*0.85, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "Black")
pif_white    <- pif(p = 0.084, p_cft = 0.084*0.85, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp, label = "White")
```

The total fraction of the overall population can be aggregated with `pif_total`:

```{r}
pif_population <- pif_total(pif_white, pif_hispanic, pif_black, 
                            pif_asian, weights = weights, var_weights = 0,
                            label = "All")
pif_population
```
We can use the `as.data.frame` function to create a `data.frame` with all of the results:

```{r}
df <- as.data.frame(pif_population, pif_white, pif_black, pif_hispanic, pif_asian)
df
```

Graphically we can visualize the fractions:


```{r}
if (requireNamespace("ggplot2", quietly = TRUE)){
  library(ggplot2)
  
  df$label <- factor(df$label, levels = df$label)
  
  ggplot(df) +
    geom_errorbar(aes(x = label, ymin = ci_low, ymax = ci_up, color = label)) +
    geom_point(aes(x = label, y = value, color = label)) +
    theme_bw() +
    labs(
      x = "Race/Ethnicity",
      y = "Potential Impact Fraction",
      title = "Prevented dementia cases",
      subtitle = "15% reduction in smoking"
    ) +
    theme(legend.position = "none") +
    scale_y_continuous(label = scales::percent_format())
}
```


Averted cases can be calculated with the same number as before with the `averted_cases` function:

```{r}
averted_cases(426.5, pif_population, variance = 2647)
```

### Example 3: A categorical population attributable fraction

[@pandeya](https://doi.org/10.1111/1753-6405.12456)
estimate the population attributable fraction of different cancers given
distinct levels of alcohol consumption for the Australian population. 

The `alcohol` dataframe included in the package contains an estimate
of the intake of alcohol by sex as well as the proportion of
individuals in each of the intake-categories:

```{r}
data(alcohol)
```

```{r, echo = FALSE}
alcohol[1:12,c("sex", "alcohol_g", "median_intake", "age_18_plus")]
```

The relative risk of alcohol varies by consumption level (`alcohol_g`). For example
in the case of rectal cancer the estimated relative risk is 1.10 (1.07–1.12) per 10g/day. We will need to compute the relative risk for each of the consumption level groups using their median intake as follows:

```{r}
#Get alcohol intake for men divided by 10 cause RR is per 10g/day
alcohol_men_intake <- c(0, 0.8, 1.6, 3.2, 7.1, 12.6, 17.4, 24.5, 34.7, 44.2, 54.4, 85.2) / 10 

#Divided by 10 cause risk is per 10 g/day
relative_risk <- exp(log(1.10)*alcohol_men_intake)

#Divide by 100 to get the prevalence in decimals
prevalence    <- c(0.472, 0.014, 0.022, 0.081, 0.081, 0.069, 0.046, 0.073, 0.042, 0.032, 0.02, 0.048)

#Calculate the paf
paf(prevalence, beta = relative_risk, var_p = 0, var_b = 0)
```

If we want to add uncertainty we can approximate the covariance of the relative risk as follows:

```{r}
#Calculate covariance of the betas given by variance*intake_i*intake_j
beta_sd <- (1.12 - 1.07) / (2*qnorm(0.975))
beta_var <- matrix(NA, ncol = length(prevalence), nrow = length(prevalence))
for (k in 1:length(prevalence)){
  for (j in 1:length(prevalence)){
    beta_var[k,j] <- beta_sd^2*(alcohol_men_intake[j])*(alcohol_men_intake[k])
  }
}

#Calculate the paf
paf_males <- paf(prevalence, beta = relative_risk, var_p = 0, var_b = beta_var, 
               label = "Males")
paf_males
```

We can repeat the same process for females:

```{r}
#Get alcohol intake for women divided by 10 cause RR is per 10g/day
alcohol_female_intake <- c(0, 0.8, 1.6, 3.9, 7.1, 12.6, 17.4, 24.5, 34.7, 44.2, 54.4, 78.1) / 10

#Divided by 10 cause risk is per 10 g/day
relative_risk <- exp(log(1.10)*alcohol_female_intake)

#Divide by 100 to get the prevalence in decimals
prevalence    <- c(0.603, 0.02, 0.054, 0.091, 0.081, 0.061, 0.026, 0.034, 0.015, 0.007, 0.003, 0.005)

#Calculate covariance of the betas given by variance*intake_i*intake_j
beta_sd <- (1.12 - 1.07) / (2*qnorm(0.975))
beta_var <- matrix(NA, ncol = length(prevalence), nrow = length(prevalence))
for (k in 1:length(prevalence)){
  for (j in 1:length(prevalence)){
    beta_var[k,j] <- beta_sd^2*(alcohol_female_intake[j])*(alcohol_female_intake[k])
  }
}

#Calculate the paf
paf_females <- paf(prevalence, beta = relative_risk, var_p = 0, var_b = beta_var, 
                   label = "Females")

paf_females
```

The total fraction can be computed with `paf_total` assuming both males and females are 50% each of the population:

```{r}
paf_total(paf_males, paf_females, weights = c(0.5, 0.5))
```

### Example 4: Combining fractions of different risks into an ensemble

@lee combines multiple population attributable fractions from different risk exposures can be combined into a single fraction. For example, we can create an attributable fraction for **behavioral** risk factors by joining the fractions for: smoking, physical inactivity, and excesive alcohol consumption into one fraction that represents the reduction in cases if all of these exposures were taken to the level of minimum risk. 

For that purpose we calculate each of the fractions separately and then use the `paf_ensemble` function to join them. We first obtain the log-relative risk, its standard deviation, and the total proportion of individuals exposed (here we highlight the part of the table):

```{r}
data(dementiarisk)
```

```{r, echo = FALSE}
dementiarisk[which(dementiarisk$risk_factor %in% 
                     c("Excessive alcohol","Smoking","Physical inactivity")),
             c("risk_factor", "logrr", "sdlog", "total")]
```

We then calculate the population attributable fraction for each:

```{r}
paf_alcohol  <- paf(p = 0.036, beta = 0.1655144, var_beta = 0.05402095^2, 
                    var_p = 0, label = "Alcohol")
paf_smoking  <- paf(p = 0.085, beta = 0.4637340, var_beta = 0.16548657^2, 
                    var_p = 0, label = "Smoking")
paf_physical <- paf(p = 0.628, beta = 0.3293037, var_beta = 0.09296182^2, 
                    var_p = 0, label = "Physical Activity")
```

The `paf_ensemble` function allows to calculate the ensemble `paf` by estimating the product:

$$
\textrm{PAF}_{\text{Ensemble}} = 1 - (1 - w_1 \cdot\textrm{PAF}_{\text{Alcohol}})\cdot (1 - w_2 \cdot\textrm{PAF}_{\text{Smoking}})\cdot (1 - w_3\cdot \textrm{PAF}_{\text{Physical}})
$$

where the weights are oftentimes calculated via a communality correction or are set to 1 for disjointed risks. In the case of this paper the communality estimates were:

+ **Physical inactivity:** 31.1

+ **Smoking:** 5.9

+ **Excessive alcohol:** 3.8

hence we can use them as the `weights` argument:

```{r}
paf_ensemble(paf_alcohol, paf_physical, paf_smoking, 
             weights = c(0.038, 0.311, 0.059),
             label = "Behavioural factors")
```

Finally, variance among the weights can be implemented with the `var_weights` argument providing a variance-covariance matrix:


```{r}
#Variance covariance matrix
weight_variance <- matrix(
  c( 1.00, 0.21, -0.02, 
     0.21, 1.00,  0.09,
    -0.02, 0.09,  1.00), byrow = TRUE, ncol = 3
)
colnames(weight_variance) <- c("Alcohol", "Physical inactivity", "Smoking")
rownames(weight_variance) <- c("Alcohol", "Physical inactivity", "Smoking")

paf_ensemble(paf_alcohol, paf_physical, paf_smoking, 
             weights     = c(0.038, 0.311, 0.059),
             var_weights = weight_variance,
             label       = "Behavioural factors")
```


### Example 6: Ensuring non-negative fractions

@lee2024comparison show an example with a relative risk of `rr = 7.0` (log relative risk `1.94591`)
and a log-variance of `0.35.` that results in a negative confidence interval when applying directly the delta method:

```{r}
paf(p = 0.05, beta = 1.94, var_beta = 0.591^2, var_p = 0)
```

The `pif` and `paf` functions contain the `logit` link option for guaranteeing positivity in the fraction's intervals: 

```{r}
paf(p = 0.05, beta = 1.94, var_beta = 0.591^2, var_p = 0, link = "logit")
```

We remark that this option is not turned on by default as the **positivity** of the fraction depends on:

1. The relative risk being strictly greater than 1. 

2. Epidemiological rationale on why the fraction should be positive. 


## References
