---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: Referencias.bib
csl: american-journal-of-epidemiology.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(deltapif)
```

To estimate a population attributable fraction two ingredients are required:

- `beta`: the relative risk or a value from which to calculate the relative risk (with variance `var_beta`).
- `p`: the exposure prevalence (with variance `var_p`).

and in the case of potential impact fractions we also require:
- `p_cft`: the counterfactual prevalence.  

> **Note** An important hypothesis of the current method is that
the relative risk estimate and the prevalence of exposure are both independent
in the sense that they were estimated in different populations and studies. 

For example consider the population attributable fraction of depression into dementia.
Consider:

-  A relative risk of  relative risk of `1.90` with a variance of `0.02`. 
-  A prevalence of depression of 12% (`0.12`) with variance `0.0009`. 

We can calculate a point estimate of the PAF as follows:

```{r}
paf(p = 0.12, beta = 1.90, var_p = 0.0009, var_beta = 0.02)
```

A potential impact fraction of reducing the prevalence of depression to 6% can be
calculated as follows: 

```{r}
pif(p = 0.12, p_cft = 0.06, beta = 1.90, var_p = 0.0009, var_beta = 0.02)
```

The following examples show how to calculate the PIF and PAF under different scenarios
as well as how to combine the fractions of subpopulations and of different risk estimates. 



## Example 1: A dichotomous potential impact fraction

The article [@lee](https://doi.org/10.1001/jamanetworkopen.2022.19672)
discusses the population attributable fraction (PAF) of dementia associated with 12 
different risk factors in US adults. It also estimates a potential impact
fraction correspondent to a 15% reduction in the exposures of such factors. 

The dataset `dementiarisk` contains the relative risk of dementia for different
exposures as well as the prevalence of exposure for the total population
and for different subpopulations (races):

```{r}
data(dementiarisk)
```

```{r, echo = FALSE}
dementiarisk[,c("risk_factor","logrr","sdlog", "total", "hispanic", "asian", "black", "white")]
```

Here we show how to calculate the potential impact fraction and the population attributable fraction for smoking among the 4 subpopulations (race groups) and then how to calculate the total attributable fraction. 

For that purpose we will choose smoking which has a log relative risk of `beta = 0.4637340` with variance of	`var_beta = 0.0273858` (`= 0.165486566^2`). 

Among hispanic individuals 6.9% (`p = 0.069`) smoke. Hence their paf is:

```{r}
paf_hispanic <- paf(p = 0.069, beta = 0.4637340, var_beta = 0.0273858, 
                    rr_link = exp)
paf_hispanic
```

Note that the model throws a warning for us not specifying the variance of `var_p`. We will 
assume that the value was perfectly estimated and it has no associated variance thus setting `var_p = 0`:

```{r}
paf_hispanic <- paf(p = 0.069, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp)
paf_hispanic
```

The `coef`, `confint`, `summary`, and `as.data.frame` functions have been extended to 
allow the user to extract information from a `paf` (or `pif`):

```{r}
as.data.frame(paf_hispanic)
```

The other fractions can be computed in the same way noting that 4.9	of non-hispanic asians, 11.7 of non-hispanic blacks and	8.4 of non-hispanic whites smoke:

```{r}
paf_asian <- paf(p = 0.049, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp)
paf_black <- paf(p = 0.117, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp)
paf_white <- paf(p = 0.084, beta = 0.4637340, var_beta = 0.0273858, 
                    var_p = 0, rr_link = exp)
```

Correlations (`correlation`) and covariances (`covariance`) can also be calculated for multiple `paf` objects:

```{r}
covariance(paf_white, paf_hispanic, paf_asian, paf_black)
```
### Combining subpopulations into a total

We can calculate the total population attributable fraction by combining the fractions of the subpopulations
weighted by the proportion of the population. According to Wikipedia the distribution in 2020 was as follows:

```{r}
pif_weights <- c("white" = 0.5784, "hispanic" = 0.1873, "black" = 0.1205, "asian" = 0.0592)

#Normalized pif_weights to sum to 1
pif_weights <- pif_weights / sum(pif_weights)
```

The `paf_total` function computes the aggregated `paf` of the whole population:

```{r}
paf_total(paf_white, paf_hispanic, paf_black, paf_asian, pif_weights = pif_weights, sigma_pif_weights = 0)
```

where we set `sigma_pif_weights = 0` as no covariance matrix was known for the race distribution data (it does exist its just not on Wikipedia). 

### Why are we using the log relative risk?

A common question is why in this case we are using the log relative risk (`logrr`) with `exponential` link
and not the relative risks. The reason depends on what standard deviation can be properly
recovered. Recall that the classic (Wald-type) confidence intervals should have the same
distance from the point estimate (in this case the `RR`) to the bounds. We can see that
this does not happen here as for example in the case of **smoking** the RR is
1.59 [1.15-2.20] and:

```{r}
2.20 - 1.59
```

while 

```{r}
1.59 - 1.15
```

The distances are not the same!

However in the log-scale they are much more similar:

```{r}
log(2.20) - log(1.59)
```

```{r}
log(1.59) - log(1.15)
```


which suggests that the uncertainty was estimated in the log-scale and hence we
need to use the log relative risk as beta and exponentiate it (`link = "exponential"`). 
Furthermore we can go to the original paper, check the methodology and conclude
that the confidence interval (and hence the variance!) was estimated using the log relative risk 
and then exponentiated. Hence  the uncertainty should be propagated for the 
log risk and then exponentiated. 

Mathematically the explanation of why this happens is because usually relative risks,
odds ratios and hazard ratios are transformed to the log-scale to estimate
the confidence intervals. You can see in Wikipedia, for example, 
the classic formulas for the [relative risks CIs](https://en.wikipedia.org/wiki/Relative_risk)
and for the [odds ratios](https://en.wikipedia.org/wiki/Odds_ratio#Statistical_inference)
all involve a log-transform. 


## Example 2: A categorical population attributable fraction

The article [@pandeya](https://doi.org/10.1111/1753-6405.12456)
estimates the population attributable fraction of different cancers given
distinct levels of alcohol consumption for the Australian population. 

The `alcohol` dataframe included in the package contains an estimate
of the intake of alcohol by sex as well as the proportion of
individuals in each of the intake-categories:

```{r}
data(alcohol)
```

```{r, echo = FALSE}
alcohol[,c("sex", "alcohol_g", "median_intake", "age_18_plus")]
```

The relative risk of alcohol varies by consumption level (`algohol_g`). For example
in the case of rectal cancer the estimated relative risk is 1.10 (1.07–1.12) per 10g/day. We will need to compute
the relative risk for each of the consumption level groups using their median intake as follows:

```{r}
#Get alcohol intake for men divided by 10 cause RR is per 10g/day
alcohol_men_intake <- c(0, 0.8, 1.6, 3.2, 7.1, 12.6, 17.4, 24.5, 34.7, 44.2, 54.4, 85.2) / 10 

#Divided by 10 cause risk is per 10 g/day
relative_risk <- exp(log(1.10)*alcohol_men_intake)

#Divide by 100 to get the prevalence in decimals
prevalence    <- c(0.472, 0.014, 0.022, 0.081, 0.081, 0.069, 0.046, 0.073, 0.042, 0.032, 0.02, 0.048)

#Calculate the paf
paf(prevalence, beta = relative_risk, var_p = 0, var_b = 0)
```

If we want to add uncertainty we can approximate the covariance of the relative risk as follows:

```{r}
#Calculate an approximate covariance of the betas this is given by var*intake*intake 
beta_sd <- (1.12 - 1.07) / (2*qnorm(0.975))
beta_var <- matrix(NA, ncol = length(prevalence), nrow = length(prevalence))
for (k in 1:length(prevalence)){
  for (j in 1:length(prevalence)){
    beta_var[k,j] <- beta_sd^2*(alcohol_men_intake[j]/10)*(alcohol_men_intake[k]/10)
  }
}


#Calculate the paf
paf(prevalence, beta = relative_risk, var_p = 0, var_b = beta_var)
```

<!--
## Example 3: Negative fractions

## Example 4: Combining fractions of different risks into an ensemble

graphPAF::paf_levin(conf_prev = 0.085, conf_RR=c(1.15, 2.20))
-->

## References
